Efficacy of COVID-19 vaccination against severe COVID-19 and risk to  double-vaccinated individuals in Scotland: relationship to co-morbidities conferring  clinically extremely vulnerable status or moderate risk

### 2021-08-29 vaxshieldupdate.Rmd
reinfections

@@

The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model (R function `survival::clogit`, package version_3.2-7, R version 3.6.3) fitted to the full dataset, specifying effects \ensuremath{\beta_{R2}, \dots, \beta_{RJ}} for the log rate ratio associated with risk categories 2 to \ensuremath{J} (\ensuremath{\beta_{R1} = 0} for the reference category \ensuremath{J=1}), and nested effects \ensuremath{\beta_{V1}, \dots, \beta_{VJ}} for the log rate ratio associated with vaccination in each of the \ensuremath{J} risk categories. With this incidence density sampling design, the conditional odds ratio is the rate ratio.  The efficacy of vaccination is 1 minus the rate ratio.  The strata in the conditional logistic regression model are matched sets: typically one case and up to 10 controls, but where two cases with the same age, sex and GP practice present on the same day the matched set comprises two cases and up to 20 controls.  When estimating the rate ratio within a risk group such as a shielding category, only cases and controls in that risk group are retained in the matched sets.   Matched sets in which all are vaccinated or none are vaccinated do not contribute to the rate ratio.   The unconditional odds ratios calculated from frequency tables of the vaccination status of cases and controls in each risk group cannot be used to estimate rate ratios [@breslow_estimation_1978;@greenland_need_1982].   The design controls not only for the matching factors of age, sex and general practice but also for calendar time.  Other variables included as covariates in the models had been previously identified as strongly associated with severe COVID-19: care home residence, number of adults in the household, number of distinct drug classes (BNF subparagraph codes) for which prescriptions were dispensed between 15 and 240 days before presentation date, and any hospital in-patient stay from 5 to 14 days before presentation date. 
<!---
# Cohort analysis of the shielding list
The case-control study can estimate only rate ratios.  To investigate the absolute rates of severe disease in those listed as clinically extremely vulnerable, we undertook a separate cohort study of all individuals who had ever been on the shielding list. At the time of this study  the dates of addition and removal from the shielding list, and linkage of the full cohort to vaccination status and mortality were not available to us, so this analysis is only provisional.  A Poisson regression model was fitted to a dataset formatted with one observation per 28-day person-time interval, and individuals censored at first event.  The event was defined as severe COVID-19 presenting within the person-time interval, and the covariates were sex, age, and shielding eligibility group.  The baseline hazard rate was specified as a natural spline function with 4 degrees of freedom.  Similar models were fitted with event indicator defined as hospitalized or fatal cases, or as any diagnosed case.  
--->

# Results
## Time course of vaccine efficacy against severe disease 
Fig \ref{fig:plotcases} shows the daily numbers of severe cases (averaged over 3-day sliding windows) by risk category.  In the first half of July the average daily number of severe cases among the 200,000 people still on the shielded list was about 4, equating to an incidence rate of about 0.6 per 1000 per month.

Fig \ref{fig:plotrateratios} shows the time course of rate ratios (computed over 42-day sliding windows) for risk category and vaccination status, in a model that includes both these variables together with care home residence and recent hospital stay as covariates.  This shows that the rate ratio for severe disease associated with shielding eligibility, adjusted for vaccination status, has increased since 1 May, even though the absolute rate of severe disease in this group remains much lower than in December-January.  The rate ratios for severe disease associated with 1 dose and 2 doses of vaccine at least 14 days earlier (with unvaccinated as reference category) declined  steadily from January to June, implying that the efficacy increased despite the rapid replacement of the Alpha variant by the Delta variant during May 2020 [@sheikh_sarscov2_2021].

@@

Table \ref{tab:rateratiosriskgroups} shows rate ratios associated with 1 and 2 vaccine doses (with 0 doses as reference category) in each of the three broad risk categories, in a model that includes risk categories (no risk conditions as reference category),  care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates. The rate ratio associated with two doses of vaccine was `r format.estci(table.riskgr[effect=="No risk condition: vax dose 2", rateratio])` in those without risk conditions, `r format.estci(table.riskgr[effect=="Moderate risk condition: vax dose 2", rateratio])` in those with moderate risk conditions, and `r format.estci(table.riskgr[effect=="Eligible for shielding: vax dose 2", rateratio])` in those eligible for shielding.  On the scale of efficacy, these estimates are equivalent to 94% in those without risk conditions, 91% in those with risk conditions, and only 68% in those eligible for shielding.  To test formally for an effect of risk category on vaccine efficacy, a model was fitted with interaction terms for  risk category \ensuremath{\times} times vaccine dose.  In those with 2 doses, the effect (on the scale of rate ratio) of eligibility for shielding (with no risk condition as reference category) on the rate ratio associated with 2 doses of vaccine (0 doses as reference category) was `r format.estci(vaxriskgr.interaction)`.

To exclude the possibility that misclassification of deaths within 28 days of a positive test as deaths from COVID-19 might account for apparently lower efficacy of vaccines in the clinically extremely vulnerable, the analysis was repeated with exclusion of fatal cases where COVID-19 was not certified as the underlying cause of death.  With this restricted case definition, the rate ratio associated with two doses of vaccine was `r format.estci(table.nonfatalorucod[effect=="No risk condition: vax dose 2", rateratio])` in those without risk conditions, `r format.estci(table.nonfatalorucod[effect=="Moderate risk condition: vax dose 2", rateratio])` in those with moderate risk conditions, and `r format.estci(table.nonfatalorucod[effect=="Eligible for shielding: vax dose 2", rateratio])` in those eligible for shielding.  
Table \ref{tab:hosprateratiosriskgroups} shows the same model with case definition broadened to include all hospitalised cases and their matched controls: the efficacy of 2 doses was lower in those eligible for shielding than in the other two risk categories.

Table \ref{tab:rateratiosshieldgroups} shows rate ratios per vaccine dose, by broad risk category and shielding eligibility group.  The efficacy per dose was lower in all shielding eligibility groups than in those without risk conditions  Table \ref{tab:hosprateratiosshieldgroups} shows the same model with case definition broadened to include all hospitalised cases and their matched controls.  Although there is evidence of efficacy in all shielding eligibility groups, the efficacy estimate was lowest in solid organ transplant recipients.

@@

# Discussion
## Statement of principal findings
1. Although the efficacy of vaccination against severe disease in those without risk conditions remains high (at least 90%), it is now clear that the efficacy in clinically extremely vulnerable individuals is considerably lower (about 70%).  There is however a dose-response effect in this group: two doses are more effective than one.

@@

The main limitation of this study is that no data on hospital prescribing are available.   Most biologic immunosuppressants and drugs for cancer are prescribed only in hospital, and the records of prescribing are not held in electronic form.  The national shielding list does not hold any information about drug therapy, and no information about diagnoses other than the seven broad categories listed as eligible for shielding.

@@

@@

p.byelig

@@

p.rateratio

>>>>

The effect of vaccination in each of the clinical vulnerability categories was estimated in a conditional logistic regression model (R function `survival::clogit`, package version_3.2-7, R version 3.6.3) fitted to the full dataset, specifying effects \ensuremath{\beta_{R2}, \dots, \beta_{RJ}} for the log rate ratio associated with risk categories 2 to \ensuremath{J} (\ensuremath{\beta_{R1} = 0} for the reference category \ensuremath{J=1}), and nested effects \ensuremath{\beta_{V1}, \dots, \beta_{VJ}} for the log rate ratio associated with vaccination in each of the \ensuremath{J} risk categories. The design controls not only for the matching factors of age, sex and general practice but also for calendar time.

@@

With this incidence density sampling design, the conditional odds ratio is the rate ratio.  The efficacy of vaccination is 1 minus the rate ratio.  The strata in the conditional logistic regression model are matched sets: typically one case and up to 10 controls, but where two cases with the same age, sex and GP practice present on the same day the matched set comprises two cases and up to 20 controls.  When estimating the rate ratio within a risk group such as a shielding category, only cases and controls in that risk group are retained in the matched sets.   Matched sets in which all are vaccinated or none are vaccinated do not contribute to the rate ratio.   The unconditional odds ratios calculated from frequency tables of the vaccination status of cases and controls in each risk group cannot be used to estimate rate ratios [@breslow_estimation_1978;@greenland_need_1982].   
# Cohort analysis of the shielding list
The case-control study can estimate only rate ratios.  To investigate the absolute rates of severe disease in those listed as clinically extremely vulnerable, we undertook a separate cohort study of all individuals who had ever been on the shielding list. At the time of this study  the dates of addition and removal from the shielding list, and linkage of the full cohort to vaccination status and mortality were not available to us, so this analysis is only provisional.  A Poisson regression model was fitted to the cohort formatted with one observation per 28-day person-time interval, and individuals censored at first diagnosis of COVID-19.  Event status was defined as severe COVID-19 presenting within the person-time interval, and the covariates were sex, baseline age, and shielding eligibility group.  The baseline hazard rate was modelled as a natural spline function of calendar time with 6 degrees of freedom, allowing the model to fit an epidemic with multiple waves.  Similar models were fitted with event status defined as hospitalized or fatal cases, or as any diagnosed case. To present the fitted hazard rates in a form that is interpretable as the risk associated with a single wave of the epidemic, the 60-day incidence rate was calculated for each time interval.

# Results
## Shielding cohort
By the date of the latest extract of the shielding list in July 2021, 8230 (4%) of the 202510 individuals who had ever been on the shielding list had been diagnosed with COVID-19, and 18893 (9%) had been removed from the list for other reasons.  Fig \ref{fig:caseseverity} shows the fitted 60-day incidence of any diagnosis and of severe cases in Poisson regression models fitted to the entire cohort.  Fig \ref{fig:plotcases} shows the daily numbers of severe cases (averaged over 3-day sliding windows) by risk category. Although the incidence of any diagnosed COVID-19 in the shielding cohort rose steeply from 1 May onwards, the incidence of severe COVID-19 remained relatively low (less than 2 per 1000 per 60 days).

<!---Fig \ref{fig:plotrateratios} shows the time course of rate ratios (computed over 42-day sliding windows) for risk category and vaccination status, in a model that includes both these variables together with care home residence and recent hospital stay as covariates.  This shows that the rate ratio for severe disease associated with shielding eligibility, adjusted for vaccination status, has increased since 1 May, even though the absolute rate of severe disease in this group remains much lower than in December-January.  The rate ratios for severe disease associated with 1 dose and 2 doses of vaccine at least 14 days earlier (with unvaccinated as reference category) declined  steadily from January to June, implying that the efficacy against severe disease increased despite the rapid replacement of the Alpha variant by the Delta variant during May 2020 [@sheikh_sarscov2_2021].--->

@@

Table \ref{tab:rateratiosriskgroups} shows rate ratios associated with 1 and 2 vaccine doses (with 0 doses as reference category) in each of the three broad risk categories, in a model that includes risk categories (no risk conditions as reference category),  care home residence, number of adults in household, number of drug classes and recent hospital stay as covariates. The rate ratio associated with two doses of vaccine was `r format.estci(table.riskgr[effect=="No risk condition: vax dose 2", rateratio])` in those without risk conditions, `r format.estci(table.riskgr[effect=="Moderate risk condition: vax dose 2", rateratio])` in those with moderate risk conditions, and `r format.estci(table.riskgr[effect=="Eligible for shielding: vax dose 2", rateratio])` in those eligible for shielding.  On the scale of efficacy, these estimates are equivalent to 94% in those without risk conditions, 91% in those with risk conditions, and only 68% in those eligible for shielding.

The estimates of efficacy for a single dose in Table \ref{tab:rateratiosriskgroups} differ from those we reported based on data up to 16 March 2021, with slightly different covariate adjustments [@mckeigue_efficacy_2021].  However when we restricted the current dataset to exclude cases and controls sampled after 16 March 2021, the point estimates of the rate ratios associated with one dose of vaccine in the same model for those with no risk conditions, moderate risk conditions and eligible for shielding were 0.28, 0.25 and 0.31 respectively, similar to the estimates reported based on data without follow-up of outcome in cases after 16 March 2021 . [@mckeigue_efficacy_2021]

To test formally for an effect of risk category on vaccine efficacy, a model was fitted with interaction terms for  risk category \ensuremath{\times} times vaccine dose.  In those with 2 doses, the effect (on the scale of rate ratio) of eligibility for shielding (with no risk condition as reference category) on the rate ratio associated with 2 doses of vaccine (0 doses as reference category) was `r format.estci(vaxriskgr.interaction)`.

The rate ratios for severe disease associated with one dose of vaccine were To exclude the possibility that misclassification of deaths within 28 days of a positive test as deaths from COVID-19 might account for apparently lower efficacy of vaccines in the clinically extremely vulnerable, the analysis was repeated with exclusion of fatal cases where COVID-19 was not certified as the underlying cause of death.  With this restricted case definition, the rate ratio associated with two doses of vaccine was `r format.estci(table.nonfatalorucod[effect=="No risk condition: vax dose 2", rateratio])` in those without risk conditions, `r format.estci(table.nonfatalorucod[effect=="Moderate risk condition: vax dose 2", rateratio])` in those with moderate risk conditions, and `r format.estci(table.nonfatalorucod[effect=="Eligible for shielding: vax dose 2", rateratio])` in those eligible for shielding.  
Table \ref{tab:hosprateratiosriskgroups} shows the same model with case definition broadened to include all hospitalised cases and their matched controls: the efficacy of 2 doses was lower in those eligible for shielding than in the other two risk categories.

Table \ref{tab:rateratiosshieldgroups} shows rate ratios per vaccine dose, by broad risk category and shielding eligibility group.  The efficacy per dose was lower in all shielding eligibility groups than in those without risk conditions  Table \ref{tab:hosprateratiosshieldgroups} shows the same model with case definition broadened to include all hospitalised cases and their matched controls.

@@

# Discussion
## Statement of principal findings
1. Although the efficacy of 2 doses of vaccine against severe disease in those without risk conditions remains high (at least 90%), it is now clear that the efficacy in clinically extremely vulnerable individuals is lower (about 70%).  Comparison between this updated report and our earlier report indicates that efficacy in the clinically extremely vulnerable is now lower than when the Alpha variant was the dominant strain in Scotland.  There is however a dose-response effect in this group as in those with no risk conditions: two doses are more effective than one.

@@

The main limitation of this study is that no data on hospital prescribing are available.   Most immunosuppressants and drugs for cancer are prescribed only in hospital, and the records of prescribing are not held in electronic form.  The national shielding list does not hold any information about drug therapy, and no information about diagnoses other than the seven broad categories listed as eligible for shielding.

@@

# Figures

```{r caseseverity, echo=FALSE, warning=FALSE, message=FALSE, fig.pos = "H", fig.width=6, fig.asp=0.8, fig.cap="\\label{fig:caseseverity} Poisson regression model for 90-day incidence of any case and severe case in the shielding cohort"}

p.anycase

@@

p.byelig + theme(legend.title = element_blank())

@@

theme_set(theme_gray(base_size = 10))
# p.rateratio
